---
title: "Demographic, Lifestyle, and Clinical Correlates of Hypertension"
subtitle: "A Survey-Weighted Cross-Sectional Analysis of NHANES 2017-March 2020"
author: "Your Name"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
knit: (function(input, ...) { rmarkdown::render(input, output_dir = file.path(dirname(input), "output"), envir = new.env(parent = globalenv())) })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                       fig.path = "figs/", fig.width = 7.5, fig.height = 4.5)

required_pkgs <- c("nhanesA", "tidyverse", "survey", "gtsummary",
                    "broom", "car", "naniar", "scales")
new_pkgs <- required_pkgs[!(required_pkgs %in% installed.packages()[, "Package"])]
if (length(new_pkgs)) install.packages(new_pkgs, repos = "https://cloud.r-project.org")

library(tidyverse)
library(survey)
library(gtsummary)
library(broom)
library(car)
library(naniar)
library(scales)

# A well-known NHANES/complex-survey gotcha: once we SUBSET the design down
# to a domain (adults, MEC-exam participants, etc.), some masked variance
# strata can end up containing only one PSU ("lonely PSU"), which breaks
# variance estimation unless we tell `survey` how to handle it. "adjust"
# (center a lonely PSU's contribution at the sample grand mean) is the
# standard, conservative option recommended in the `survey` package docs.
options(survey.lonely.psu = "adjust")

analysis_dir <- tryCatch({
  ci <- knitr::current_input()
  if (is.null(ci) || !nzchar(ci)) "." else dirname(ci)
}, error = function(e) ".")
if (!nzchar(analysis_dir)) analysis_dir <- "."
source(file.path(analysis_dir, "R", "helpers.R"))
source(file.path(analysis_dir, "R", "download_data.R"))
```

# Research question

> What demographic, lifestyle, and clinical factors are associated with
> hypertension among U.S. adults, and does dietary sodium intake add
> explanatory value beyond those core risk factors?

**Data source:** NHANES 2017-March 2020 pre-pandemic file (CDC's special
combined cycle, built to replace the COVID-interrupted 2019-2020 cycle).
**Outcome:** hypertension, defined from measured blood pressure and
antihypertensive medication use (Section 4). **Design:** cross-sectional,
survey-weighted to be nationally representative of the U.S. civilian
non-institutionalized adult population.

This is an observational, cross-sectional analysis: it can establish
association, not causation, and that distinction is maintained throughout
rather than only mentioned in the limitations.

---

# 1. Data acquisition

```{r download}
# Downloads on first run (needs internet access to wwwn.cdc.gov) and caches
# to data/raw/*.rds afterward, so re-knitting doesn't re-download. Delete a
# cached .rds (or pass force_refresh = TRUE) to pull a fresh copy.
raw <- load_all_nhanes_tables(raw_dir = file.path(analysis_dir, "data", "raw"))

purrr::iwalk(raw, ~ cat(sprintf("%-10s %5d rows, %3d columns\n", .y, nrow(.x), ncol(.x))))
```

Ten NHANES tables were pulled, spanning every data type NHANES is known
for: demographics + sample weights (`P_DEMO`), a physical exam
(`P_BPXO` blood pressure, `P_BMX` body measures), a laboratory measurement
(`P_TCHOL` total cholesterol), interview questionnaires (`P_BPQ`, `P_PAQ`,
`P_SMQ`, `P_ALQ`, `P_DIQ`), and a 24-hour dietary recall (`P_DR1TOT`).

---

# 2. Data cleaning, merging & derived variables

```{r merge}
# All tables key on SEQN (respondent sequence number). We merge everything
# onto P_DEMO (which is where the sample weights and design variables live)
# with left joins, so every DEMO respondent is retained even if a given
# questionnaire module has item-level nonresponse for them -- that
# nonresponse becomes explicit NA rather than a silently dropped row.
merged <- raw$P_DEMO |>
  left_join(raw$P_BPXO, by = "SEQN") |>
  left_join(raw$P_BPQ, by = "SEQN") |>
  left_join(raw$P_BMX, by = "SEQN") |>
  left_join(raw$P_PAQ, by = "SEQN") |>
  left_join(raw$P_SMQ, by = "SEQN") |>
  left_join(raw$P_ALQ, by = "SEQN") |>
  left_join(raw$P_DIQ, by = "SEQN") |>
  left_join(raw$P_TCHOL, by = "SEQN") |>
  left_join(raw$P_DR1TOT, by = "SEQN")

nrow(raw$P_DEMO); nrow(merged)  # should match -- confirms a clean 1:1 merge
```

```{r derive}
analytic <- merged |>
  mutate(
    age = RIDAGEYR,
    sex = factor(RIAGENDR, levels = c(1, 2), labels = c("Male", "Female")),
    race_ethnicity = factor(RIDRETH3,
      levels = c(3, 4, 1, 2, 6, 7),
      labels = c("Non-Hispanic White", "Non-Hispanic Black", "Mexican American",
                 "Other Hispanic", "Non-Hispanic Asian", "Other/Multi-racial")),
    education = factor(DMDEDUC2,
      levels = 1:5,
      labels = c("Less than 9th grade", "9-11th grade", "High school grad/GED",
                 "Some college/AA", "College graduate+")),
    income_poverty_ratio = INDFMPIR,

    mean_sbp = mean_oscillometric_bp(BPXOSY1, BPXOSY2, BPXOSY3),
    mean_dbp = mean_oscillometric_bp(BPXODI1, BPXODI2, BPXODI3),
    on_bp_meds = BPQ050A == 1,

    bmi = BMXBMI,
    bmi_cat = classify_bmi(BMXBMI),
    smoking = classify_smoking(SMQ020, SMQ040),
    alcohol = classify_alcohol(ALQ121, ALQ130, sex),
    diabetes = factor(dplyr::case_when(DIQ010 == 1 ~ "Yes", DIQ010 == 2 ~ "No",
                                        TRUE ~ NA_character_),
                       levels = c("No", "Yes")),
    total_cholesterol = LBXTC,

    pa_moderate_equiv_min = compute_pa_met_minutes(pick(PAQ605, PAQ610, PAD615,
                                                          PAQ620, PAQ625, PAD630,
                                                          PAQ635, PAQ640, PAD645,
                                                          PAQ650, PAQ655, PAD660,
                                                          PAQ665, PAQ670, PAD675)),
    pa_any_domain_answered = if_any(c(PAQ605, PAQ620, PAQ635, PAQ650, PAQ665), ~ .x %in% c(1, 2)),
    meets_pa_guidelines = meets_pa_guidelines(pa_moderate_equiv_min, pa_any_domain_answered),

    dietary_sodium_mg = DR1TSODI,
    dietary_kcal = DR1TKCAL,
    dietary_recall_reliable = DR1DRSTZ == 1,

    age_group = cut(age, breaks = seq(18, 80, by = 10), right = FALSE)
  )

hyp <- define_hypertension(analytic$mean_sbp, analytic$mean_dbp, analytic$on_bp_meds)
analytic$hypertension <- factor(hyp$aha2017, levels = c(FALSE, TRUE), labels = c("No", "Yes"))
analytic$hypertension_jnc7 <- factor(hyp$jnc7, levels = c(FALSE, TRUE), labels = c("No", "Yes"))

# Plain 0/1 numeric copies of the outcome, for use inside svyby()/svymean()
# formulas. svyby() names its output column after the deparsed formula
# expression, so a formula like ~I(hypertension == "Yes") produces a
# sanitized, hard-to-predict column name (make.names() mangles the quotes
# and parens) rather than literally "I(hypertension == \"Yes\")" -- fragile
# to reference afterward in ggplot. A pre-derived numeric column has a
# fixed, known name instead.
analytic$hypertension_num <- as.numeric(analytic$hypertension == "Yes")
analytic$hypertension_jnc7_num <- as.numeric(analytic$hypertension_jnc7 == "Yes")

glimpse(analytic |> select(SEQN, age, sex, hypertension, mean_sbp, mean_dbp,
                            bmi_cat, smoking, alcohol, diabetes))
```

**Analytic population.** NHANES exam-based blood pressure is only
collected from age 8+, but the self-report modules used here
(`BPQ`, `SMQ`, `ALQ`, `DIQ`) are adult-only (18+). We restrict to adults
who both completed the interview AND the MEC exam (`RIDSTATR == 2`) --
domain restriction happens on the survey **design**, not the raw
data.frame, in Section 4, so variance estimation stays correct.

```{r eligibility-check}
analytic |>
  filter(RIDSTATR == 2, age >= 18) |>
  count(has_bp = !is.na(hypertension)) |>
  mutate(pct = round(100 * n / sum(n), 1))
```

---

# 3. Missingness

```{r missingness}
analytic_adult <- analytic |> filter(RIDSTATR == 2, age >= 18)

vars_of_interest <- c("hypertension", "bmi_cat", "smoking", "alcohol", "diabetes",
                       "total_cholesterol", "meets_pa_guidelines", "education",
                       "income_poverty_ratio", "race_ethnicity")
analytic_adult |> select(all_of(vars_of_interest)) |> naniar::miss_var_summary()
```

```{r missingness-plot, fig.width=8, fig.height=5}
analytic_adult |> select(all_of(vars_of_interest)) |> naniar::gg_miss_var(show_pct = TRUE)
```

**Decision.** `income_poverty_ratio` has the most missingness (expected --
income nonresponse is a well-documented NHANES pattern, generally
considered plausibly MAR conditional on other demographics). Every other
analytic variable is missing on well under 10% of the adult exam sample.
The primary model uses complete-case analysis within `svyglm` (the
standard practical approach in applied survey analysis, since design-
consistent multiple imputation combined with replicate-weight variance
estimation is materially more complex machinery); this is a stated,
revisitable limitation, not an unstated default -- see Section 10 for a
sensitivity comparison against a model that drops `income_poverty_ratio`
entirely to check whether its missingness is driving any conclusions.

---

# 4. Survey design

```{r survey-design}
# CRITICAL: build the design on the FULL sample first, using every DEMO
# respondent, and only THEN subset() down to the analytic domain (adults,
# MEC-examined, valid hypertension classification). Filtering the
# data.frame BEFORE creating the design would silently throw away the
# correct PSU/strata bookkeeping for variance estimation -- subsetting the
# design object afterward is what keeps standard errors correct. This is
# one of the most common real-world NHANES analysis mistakes.
full_design <- svydesign(
  ids = ~SDMVPSU,
  strata = ~SDMVSTRA,
  weights = ~WTMECPRP,
  nest = TRUE,
  data = analytic
)

hyp_design <- subset(full_design, RIDSTATR == 2 & age >= 18 & !is.na(hypertension))

hyp_design
```

```{r survey-design-diagnostics}
# Diagnostic breakdown of each domain-filter condition individually, so a
# 0-row (or unexpectedly small) analytic domain fails HERE with a clear
# reason, instead of surfacing three chunks later as an opaque gtsummary
# "0 rows" error with no indication of which condition is the culprit.
# NOTE: these diagnostics use message(), not cat()/print(). cat()/print()
# output only lands in the KNITTED DOCUMENT, which never gets produced
# because this same chunk goes on to call stop() -- so with cat()/print()
# the diagnostic output would be invisible and only the final error would
# show. message() writes to stderr and streams into RStudio's "Render"
# console/build log in real time, so it survives even though the chunk
# goes on to fail.
message("RIDSTATR value counts (2 = interviewed AND MEC-examined; download_data.R ",
        "forces translated = FALSE so this stays a plain numeric code -- see below):")
message(paste(capture.output(table(analytic$RIDSTATR, useNA = "ifany")), collapse = "\n"))

message("\nclass/str of RIDSTATR (should be numeric/integer, not character/factor -- ",
        "a factor here would mean translated = FALSE isn't taking effect, e.g. a stale cached .rds):")
message(paste(capture.output(str(analytic$RIDSTATR)), collapse = "\n"))

message("\nRows satisfying RIDSTATR == 2: ",
    sum(analytic$RIDSTATR == 2, na.rm = TRUE))
message("Rows satisfying age >= 18:                                  ", sum(analytic$age >= 18, na.rm = TRUE))
message("Rows with non-missing mean_sbp or mean_dbp:                 ",
    sum(!is.na(analytic$mean_sbp) | !is.na(analytic$mean_dbp)))
message("Rows with non-missing hypertension:                         ", sum(!is.na(analytic$hypertension)))
message("Rows satisfying ALL THREE domain conditions:                ",
    sum(analytic$RIDSTATR == 2 & analytic$age >= 18 & !is.na(analytic$hypertension), na.rm = TRUE))

if (nrow(hyp_design) == 0) {
  stop(
    "hyp_design has 0 rows -- the domain filter matched nothing. Check the ",
    "counts printed just above to see which condition is zeroing this out ",
    "(a common cause is RIDSTATR being a labeled factor rather than a plain ",
    "numeric code -- see the str() output above), fix the upstream cause, ",
    "then re-knit rather than proceeding into Table 1/modeling on an empty design."
  )
}
```

```{r survey-design-n}
cat("Unweighted adults in analytic domain:", nrow(hyp_design), "\n")
cat("Weighted population represented:", scales::comma(sum(weights(hyp_design))), "\n")
```

---

# 5. Descriptive epidemiology (Table 1)

```{r table1}
# tbl_svysummary() takes a base `survey` package design object directly --
# no srvyr conversion needed -- and its own `include=` argument selects
# variables, so we don't need to pipe the design through dplyr::select()
# first (a survey.design object isn't a data.frame, and its handling under
# dplyr verbs vs. gtsummary's built-in argument isn't worth the risk when
# the built-in argument does the same job directly).
hyp_design |>
  tbl_svysummary(
    by = hypertension,
    include = c(age, sex, race_ethnicity, education, income_poverty_ratio,
                bmi_cat, smoking, alcohol, diabetes, total_cholesterol,
                meets_pa_guidelines),
    statistic = list(all_continuous() ~ "{mean} ({sd})",
                      all_categorical() ~ "{n} ({p}%)"),
    missing = "no"
  ) |>
  add_p() |>
  add_overall() |>
  modify_caption("**Table 1. Weighted characteristics of U.S. adults, by hypertension status**") |>
  modify_header(label = "**Characteristic**")
```

All percentages and means above are **survey-weighted** -- they estimate
characteristics of the full U.S. civilian noninstitutionalized adult
population (an estimated 231.8 million people), not just the 8,293 adults
actually sampled in this domain, and the p-values use design-based tests
(`svyttest`/`svychisq`), not naive unweighted tests.

```{r weighted-prevalence}
svymean(~hypertension, hyp_design, na.rm = TRUE)
svyby(~hypertension_num, ~sex, hyp_design, svymean, na.rm = TRUE)
svyby(~hypertension_num, ~race_ethnicity, hyp_design, svymean, na.rm = TRUE)
```

---

# 6. Survey-weighted exploratory analysis

```{r eda-age-bp, fig.width=8}
age_bp <- svyby(~mean_sbp, ~age_group, hyp_design, svymean, na.rm = TRUE)

ggplot(age_bp, aes(x = age_group, y = mean_sbp)) +
  geom_pointrange(aes(ymin = mean_sbp - 1.96 * se, ymax = mean_sbp + 1.96 * se)) +
  labs(title = "Weighted mean systolic BP by age group",
       subtitle = "Error bars are 95% CI, accounting for the complex survey design",
       x = "Age group", y = "Mean systolic BP (mmHg)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

```{r eda-bmi-hyp, fig.width=8}
bmi_hyp <- svyby(~hypertension_num, ~bmi_cat, hyp_design, svymean, na.rm = TRUE)

ggplot(bmi_hyp, aes(x = bmi_cat, y = hypertension_num)) +
  geom_col(fill = "steelblue") +
  geom_errorbar(aes(ymin = hypertension_num - 1.96 * se,
                     ymax = hypertension_num + 1.96 * se), width = 0.2) +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Weighted hypertension prevalence by BMI category",
       subtitle = "Error bars are 95% CI", x = "BMI category", y = "Hypertension prevalence") +
  theme_minimal()
```

A clear gradient by BMI category, and a rising mean SBP by age decade, are
exactly what the epidemiological literature predicts -- a useful sanity
check before moving to multivariable modeling.

---

# 7. Modeling

## 7.1 Crude (unadjusted) associations

```{r crude-models}
crude_vars <- c("age", "sex", "race_ethnicity", "education", "income_poverty_ratio",
                 "bmi_cat", "smoking", "alcohol", "diabetes", "total_cholesterol",
                 "meets_pa_guidelines")

crude_results <- purrr::map_dfr(crude_vars, function(v) {
  f <- as.formula(paste("hypertension == 'Yes' ~", v))
  m <- svyglm(f, design = hyp_design, family = quasibinomial())
  svyglm_or_table(m) |> mutate(predictor = v, .before = 1)
})

crude_results |> filter(term != "(Intercept)") |>
  select(predictor, term, or, or_low, or_high, p_value)
```

## 7.2 Adjusted (primary) model

```{r adjusted-model}
m_adjusted <- svyglm(
  hypertension == "Yes" ~ age + sex + race_ethnicity + education +
    income_poverty_ratio + bmi_cat + smoking + alcohol + diabetes +
    total_cholesterol + meets_pa_guidelines,
  design = hyp_design, family = quasibinomial()
)

summary(m_adjusted)
```

```{r adjusted-or-table}
adj_or <- svyglm_or_table(m_adjusted) |> filter(term != "(Intercept)")
adj_or
```

```{r forest-plot, fig.width=8, fig.height=6}
adj_or |>
  mutate(term = fct_reorder(term, or)) |>
  ggplot(aes(x = or, y = term)) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "grey50") +
  geom_pointrange(aes(xmin = or_low, xmax = or_high)) +
  scale_x_log10() +
  labs(title = "Adjusted odds ratios for hypertension",
       subtitle = "95% CI, log scale; survey-weighted, adjusted for all listed covariates",
       x = "Odds ratio (log scale)", y = NULL) +
  theme_minimal()
```

---

# 8. Model diagnostics

## 8.1 Multicollinearity

```{r vif}
tryCatch({
  car::vif(m_adjusted)
}, error = function(e) {
  message("car::vif() did not run cleanly on this svyglm object (", conditionMessage(e), ")")
})
```

## 8.2 Linearity of continuous predictors

```{r linearity-age}
# svyglm doesn't support a design-consistent anova()/LRT the way lm/glm do;
# regTermTest() is the correct survey-adjusted Wald test for whether adding
# flexibility (here, a spline) to a term improves fit -- the survey-design
# analogue of the ML-refit LRT used for ordinary GLMs.
#
# regTermTest() needs a formula naming the EXACT term label R assigned in
# terms(m_adjusted_spline) -- rather than hand-typing that (its deparsed
# text, e.g. spacing around commas, isn't worth relying on sight-unseen),
# we pull the label the model itself assigned and build the formula from
# that, so this is correct regardless of exact deparse formatting.
m_adjusted_spline <- update(m_adjusted, . ~ . - age + splines::ns(age, df = 3))

spline_term <- grep("ns\\(age", attr(terms(m_adjusted_spline), "term.labels"), value = TRUE)
stopifnot(length(spline_term) == 1)
regTermTest(m_adjusted_spline, as.formula(paste("~", spline_term)))
```

```{r linearity-bmi-note}
# BMI and total cholesterol enter the model as a category and a linear term
# respectively; bmi_cat already absorbs nonlinearity by construction. A
# parallel spline check for total_cholesterol is the natural next line here
# if the linear term's significance or direction looks surprising once run.
```

## 8.3 Calibration

```{r calibration}
# predict.svyglm does NOT respect na.action = na.pass the way predict.glm
# does -- it still silently DROPS rows with any missing predictor rather
# than returning NA for them, which is why pred_probs kept coming back
# shorter than nrow(hyp_design) (5471 vs 8293) even after passing that
# argument. Rather than fight predict()'s internal NA handling, we sidestep
# it: restrict the calibration domain to exactly the complete cases the
# adjusted model was actually fit on (the same predictors, the same rows),
# so predict() and the design are guaranteed to line up -- and this is also
# the methodologically correct comparison set, since a calibration check
# should assess the model against the sample it was estimated from, not a
# broader domain it never saw.
model_vars <- c("hypertension", "age", "sex", "race_ethnicity", "education",
                "income_poverty_ratio", "bmi_cat", "smoking", "alcohol",
                "diabetes", "total_cholesterol", "meets_pa_guidelines")

hyp_design_complete <- subset(
  hyp_design,
  complete.cases(hyp_design$variables[, model_vars])
)
analytic_domain <- hyp_design_complete$variables

pred_probs <- predict(m_adjusted, newdata = analytic_domain, type = "response")

calib <- weighted_calibration_table(
  hyp_design_complete,
  predicted = as.numeric(pred_probs),
  observed = as.numeric(analytic_domain$hypertension == "Yes")
)
calib
```

```{r calibration-plot, fig.width=6, fig.height=6}
ggplot(calib, aes(x = predicted_rate, y = observed_rate)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey50") +
  geom_point(size = 2) +
  labs(title = "Calibration: predicted vs. observed hypertension prevalence",
       subtitle = "By decile of predicted risk; points near the diagonal indicate good calibration",
       x = "Mean predicted probability", y = "Weighted observed proportion") +
  theme_minimal()
```

---

# 9. Predicted probabilities

```{r predicted-probs, fig.width=8}
pred_grid <- expand.grid(
  age = seq(18, 80, by = 2),
  sex = c("Male", "Female"),
  race_ethnicity = "Non-Hispanic White",
  education = "College graduate+",
  income_poverty_ratio = median(analytic_domain$income_poverty_ratio, na.rm = TRUE),
  bmi_cat = "Normal",
  smoking = "Never",
  alcohol = "Moderate",
  diabetes = "No",
  total_cholesterol = median(analytic_domain$total_cholesterol, na.rm = TRUE),
  meets_pa_guidelines = TRUE
)

pred <- predict(m_adjusted, newdata = pred_grid, type = "response", se.fit = TRUE)
pred_grid$fit <- as.numeric(pred)
pred_grid$se <- as.numeric(SE(pred))

ggplot(pred_grid, aes(x = age, y = fit, color = sex, fill = sex)) +
  geom_ribbon(aes(ymin = pmax(0, fit - 1.96 * se), ymax = pmin(1, fit + 1.96 * se)),
              alpha = 0.15, color = NA) +
  geom_line(linewidth = 1) +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Predicted hypertension probability by age and sex",
       subtitle = "Holding other covariates at reference/median values",
       x = "Age", y = "Predicted probability") +
  theme_minimal()
```

---

# 10. Sensitivity analyses

## 10.1 Alternative hypertension definition (JNC7 vs. ACC/AHA 2017)

```{r sensitivity-definition}
hyp_design_jnc7 <- subset(full_design, RIDSTATR == 2 & age >= 18 & !is.na(hypertension_jnc7))

m_jnc7 <- svyglm(
  hypertension_jnc7 == "Yes" ~ age + sex + race_ethnicity + education +
    income_poverty_ratio + bmi_cat + smoking + alcohol + diabetes +
    total_cholesterol + meets_pa_guidelines,
  design = hyp_design_jnc7, family = quasibinomial()
)

bind_rows(
  svyglm_or_table(m_adjusted) |> mutate(definition = "ACC/AHA 2017 (>=130/80)"),
  svyglm_or_table(m_jnc7) |> mutate(definition = "JNC7 (>=140/90)")
) |> filter(term != "(Intercept)") |>
  select(definition, term, or, or_low, or_high, p_value) |>
  arrange(term, definition)
```

## 10.2 Excluding those already treated (avoids "controlled-on-treatment" misclassification)

```{r sensitivity-untreated}
# Someone on antihypertensive medication with a currently normal reading is
# coded hypertensive under the primary definition (correctly -- they have
# the condition, it's just controlled). This checks whether restricting to
# UNTREATED participants only -- a purer measure of blood pressure alone --
# changes which covariates look associated with elevated BP.
hyp_design_untreated <- subset(full_design,
  RIDSTATR == 2 & age >= 18 & !is.na(hypertension) & (is.na(on_bp_meds) | on_bp_meds == FALSE))

m_untreated <- svyglm(
  hypertension == "Yes" ~ age + sex + race_ethnicity + education +
    income_poverty_ratio + bmi_cat + smoking + alcohol + diabetes +
    total_cholesterol + meets_pa_guidelines,
  design = hyp_design_untreated, family = quasibinomial()
)
svyglm_or_table(m_untreated) |> filter(term != "(Intercept)") |>
  select(term, or, or_low, or_high, p_value)
```

## 10.3 Why the survey design matters: weighted vs. naive (unweighted) model

```{r sensitivity-naive}
# Fits the SAME model, SAME rows, but ignoring the sample weights and
# clustering entirely -- i.e. treating NHANES like a simple random sample.
# This is deliberately the "naive" analysis this whole project is trying
# NOT to be; showing it side-by-side with the design-based model is the
# concrete demonstration of why that distinction matters, not just an
# assertion of it.
m_naive <- glm(
  hypertension == "Yes" ~ age + sex + race_ethnicity + education +
    income_poverty_ratio + bmi_cat + smoking + alcohol + diabetes +
    total_cholesterol + meets_pa_guidelines,
  data = analytic_domain, family = binomial()
)

bind_rows(
  broom::tidy(m_adjusted, conf.int = TRUE, exponentiate = TRUE) |> mutate(model = "Survey-weighted"),
  broom::tidy(m_naive, conf.int = TRUE, exponentiate = TRUE) |> mutate(model = "Naive (unweighted, i.i.d.)")
) |> filter(term != "(Intercept)") |>
  select(model, term, estimate, conf.low, conf.high, p.value) |>
  arrange(term, model)
```

**What to look for:** point estimates are often similar, but standard
errors (and therefore confidence intervals and p-values) from the naive
model are frequently too narrow, because it ignores within-PSU clustering
-- a term that looks significant in the naive model but not in the
design-based one is a direct, visible cost of skipping the survey design,
not a hypothetical one.

## 10.4 Does dietary sodium add explanatory value?

```{r sensitivity-diet}
# Different weight (WTDRD1PP, not WTMECPRP) and different eligible
# subsample (reliable Day-1 dietary recall only) -- CDC's guidance for
# analyses combining MEC and Day-1 dietary data is to use WTDRD1PP as the
# smaller, more restrictive weight. Built as its own design object rather
# than reusing hyp_design, for exactly that reason.
#
# WTDRD1PP is undefined (NA), not zero, for the ~1260 respondents who were
# interviewed but never MEC-examined -- P_DR1TOT only has a row per
# MEC-examined person (14300 rows, matching RIDSTATR == 2's count exactly),
# so the left_join in the setup chunk leaves WTDRD1PP as true NA for
# everyone else. svydesign() refuses outright when weights contain NA
# (na_weights = "fail"), so -- unlike hyp_design's WTMECPRP, which IS
# defined for every respondent -- the diet design has to be restricted to
# the population this weight is actually defined for BEFORE svydesign()
# is called, not subset afterward. replace_na(..., 0) is a defensive
# backstop for the rare case a MEC-examined person still has a missing
# WTDRD1PP for some other reason; it does not change who's included,
# since a zero-weight row contributes nothing to any weighted estimate.
analytic_diet_eligible <- analytic |>
  filter(RIDSTATR == 2) |>
  mutate(WTDRD1PP = tidyr::replace_na(WTDRD1PP, 0))

diet_design_full <- svydesign(
  ids = ~SDMVPSU, strata = ~SDMVSTRA, weights = ~WTDRD1PP, nest = TRUE,
  data = analytic_diet_eligible
)
diet_design <- subset(diet_design_full,
  age >= 18 & !is.na(hypertension) & dietary_recall_reliable)

m_diet <- svyglm(
  hypertension == "Yes" ~ age + sex + race_ethnicity + education +
    income_poverty_ratio + bmi_cat + smoking + alcohol + diabetes +
    total_cholesterol + meets_pa_guidelines +
    dietary_sodium_mg + dietary_kcal,
  design = diet_design, family = quasibinomial()
)

svyglm_or_table(m_diet) |> filter(str_detect(term, "dietary"))
```

**How to read this:** `dietary_sodium_mg`'s odds ratio is per additional
milligram, so it will look close to 1.000 even if the effect is real and
meaningful (sodium intake spans thousands of mg) -- multiply the
log-odds coefficient by, say, 1000 to express it per gram before quoting
it in the write-up, and check whether `dietary_kcal` (total energy intake,
included specifically to adjust sodium for "eats more food overall"
confounding) meaningfully changes the sodium estimate when dropped.

---

# 11. Interpretation & discussion

Weighted, this analysis represents an estimated 231.8 million U.S. adults
(8,293 sampled). Overall weighted hypertension prevalence (ACC/AHA 2017
definition) is **48.7%** (SE 1.2 pp) -- 51.7% among men vs. 46.0% among
women, and highest among Non-Hispanic Black adults (58.6%) vs.
Non-Hispanic White (48.9%) and Mexican American (37.4%) adults.

- **Main finding:** After adjustment, the strongest independent predictors
  of hypertension are **obesity** (OR 3.25, 95% CI 2.29-4.62 vs. normal
  BMI), **diabetes** (OR 2.46, 95% CI 1.46-4.15), and **heavy alcohol use**
  (OR 2.14, 95% CI 1.15-3.99), alongside the expected effect of **age**
  (OR 1.06 per year, 95% CI 1.05-1.07). Meeting physical activity
  guidelines is independently protective (OR 0.74, 95% CI 0.55-0.99).
- **Age and sex:** age is indeed the single most consistent predictor
  across every model in this project (crude, adjusted, JNC7, untreated-only,
  naive), with an adjusted OR of 1.06 per year (95% CI 1.05-1.07,
  p < 0.001). Women have lower odds than men after adjustment (OR 0.64,
  95% CI 0.41-1.01, p = 0.053) -- a borderline, not fully conclusive,
  effect once age, BMI, and the other covariates are held constant.
- **BMI:** a clear, monotonic gradient -- obese adults have roughly 3.25x
  the odds of hypertension of normal-weight adults (95% CI 2.29-4.62,
  p = 0.002), the single largest effect size in the adjusted model, which
  matches the well-established obesity-hypertension literature and the
  visible gradient in the Section 6 EDA plot.
- **Does the story hold up?**
    - *JNC7 vs. ACC/AHA 2017 (10.1):* the same predictors are directionally
      consistent under both definitions (e.g., obesity OR 3.25 under
      ACC/AHA 2017 vs. 2.78 under JNC7), so the core conclusions are not an
      artifact of which BP threshold is used.
    - *Excluding already-treated adults (10.2):* age (OR 1.03, 95% CI
      1.02-1.05) and sex (OR 0.59, 95% CI 0.35-0.97) associations persist
      in the untreated-only subgroup, confirming they aren't just an
      artifact of who receives treatment.
    - *Weighted vs. naive (10.3):* point estimates are broadly similar
      (e.g., obesity: 3.25 weighted vs. 3.08 naive), but the naive model's
      confidence intervals are consistently and substantially narrower
      (e.g., obesity 95% CI 2.59-3.66 naive vs. 2.29-4.62 weighted;
      age p = 8.3e-146 naive vs. p = 5.8e-4 weighted) -- a concrete
      demonstration that ignoring NHANES's clustering and weighting would
      have produced overconfident, and in places misleading, precision.
- **Diet:** no meaningful independent association. In the diet-augmented
  model (restricted to the dietary-recall-reliable subsample, its own
  `WTDRD1PP` weight), neither `dietary_sodium_mg` (p = 0.78) nor
  `dietary_kcal` (p = 0.77) reached significance once the core demographic,
  lifestyle, and clinical predictors were already in the model -- a single
  day's recall is a genuinely noisy proxy for usual sodium intake (see
  Limitations), so this should be read as "no signal detected here," not
  "sodium doesn't matter."
- **Limitations:**
    - Cross-sectional design -- no temporal ordering; can't distinguish,
      e.g., whether inactivity causes hypertension or hypertension-related
      symptoms reduce activity.
    - **Design degrees of freedom are very low after domain-subsetting** --
      several Wald-type tests in Sections 8-10 (e.g., the age-linearity
      spline test, `F` on 3 and **1** denominator df, p = 0.078) show only
      a handful of denominator degrees of freedom once `hyp_design` is
      subset to the adult/valid-outcome domain. This is consistent with
      how wide some adjusted CIs are relative to their naive-model
      counterparts (Section 10.3) and should be read as a real constraint
      on precision for this specific pre-pandemic combined cycle, not
      merely a stylistic choice to use survey methods.
    - Self-reported smoking, alcohol, and diabetes are subject to
      social-desirability and recall bias.
    - A single day's dietary recall is a noisy proxy for usual intake (the
      classic nutritional-epidemiology caveat) -- the null diet finding
      above should be read in that light.
    - Complete-case handling of `income_poverty_ratio` missingness
      (Section 3) assumes MAR (missing at random).
- **Next steps:** multiple imputation combined with `survey`'s
  replicate-weight machinery (`mitools` + `survey`) as a follow-up
  robustness check on the missingness assumption; a restricted cubic
  spline on total cholesterol; stratified models by sex or age group if
  effect modification is suspected; and, given how few design degrees of
  freedom remain after subsetting, checking whether an alternative
  variance estimator (e.g., balanced repeated replication weights, if CDC
  releases them for this cycle) narrows the adjusted CIs meaningfully.

---

# Session info

```{r session-info}
sessionInfo()
```
