---
title: "Longitudinal Depression Trajectories After Detox"
subtitle: "A Repeated-Measures / Mixed-Effects Analysis of the HELP Study"
author: "Your Name"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
# Forces the RStudio "Knit" button (and knitr::knit/rmarkdown::render run
# without an explicit output_dir) to write the rendered HTML into output/
# instead of next to this .Rmd. Without this, "Knit" writes analysis.html
# into the project root, which then drifts out of sync with output/analysis.html
# -- this makes output/analysis.html the one canonical rendered copy.
# NOTE: the "knit:" value below must stay on a single line -- YAML front
# matter does not support an unquoted multi-line flow scalar, which is what
# broke the earlier version of this field.
knit: (function(input, ...) { rmarkdown::render(input, output_dir = file.path(dirname(input), "output"), envir = new.env(parent = globalenv())) })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                       fig.path = "figs/", fig.width = 7, fig.height = 4.5)

# ---- Packages ----
required_pkgs <- c("mosaicData", "tidyverse", "naniar", "lme4", "lmerTest",
                    "nlme", "broom.mixed", "performance", "DHARMa",
                    "mice", "ggeffects", "influence.ME", "see",
                    "RLRsim", "gtsummary", "boot")
new_pkgs <- required_pkgs[!(required_pkgs %in% installed.packages()[, "Package"])]
if (length(new_pkgs)) {
  install.packages(new_pkgs, repos = "https://cloud.r-project.org")
}

library(mosaicData)
library(tidyverse)
library(naniar)
library(lme4)
library(lmerTest)   # adds p-values to lmer via Satterthwaite
library(nlme)
library(broom.mixed)
library(performance)
library(DHARMa)
library(mice)
library(ggeffects)
library(influence.ME)
library(RLRsim)      # exact finite-sample test for a single variance component
library(gtsummary)   # baseline "Table 1"
library(boot)         # backs lme4::confint(..., method = "boot")

analysis_dir <- tryCatch({
  ci <- knitr::current_input()
  if (is.null(ci) || !nzchar(ci)) "." else dirname(ci)
}, error = function(e) ".")
if (!nzchar(analysis_dir)) analysis_dir <- "."
source(file.path(analysis_dir, "R", "helpers.R"))
```

# Research question

> How does depressive symptom severity (CESD) change over 24 months
> following detoxification, and are individual trajectories associated with
> baseline substance type, housing status, and treatment assignment?

**Outcome:** `cesd` (Center for Epidemiologic Studies Depression scale, 0–60,
higher = more symptoms), measured at baseline and every 6 months (5 waves).

**Predictors of interest:** `time`, `substance`, `homeless`, `treat`, and
`age`. The supplied `HELPfull` data does not include a sex variable.

---

# 1. Data loading, cleaning & validation

```{r load-data}
data(HELPfull)

# NOTE on baseline-only covariates: in HELPfull, PRIMSUB2 (substance) and AGE
# are recorded ONLY at the baseline interview (TIME == 0) and are coded NA at
# every follow-up wave -- they are respondent characteristics collected once,
# not repeated-measures items. GROUP (treatment arm) is stored on every row
# already. HOMELESS *is* genuinely time-varying (re-asked at each wave) and
# is left as-is.
#
# Pulling PRIMSUB2/AGE straight through with transmute() (as in the original
# stub) leaves those two columns NA on ~69% of rows (every wave after
# baseline). Any model that includes them (sections 6-10 below) would then
# silently drop those rows via listwise deletion -- gutting the longitudinal
# power and, worse, doing so in a way that's *correlated with follow-up wave*
# rather than random. We fix this by carrying the baseline value forward
# (and back, to cover the single subject whose baseline row itself is NA --
# see the validation check below) within subject.

help_df <- HELPfull %>%
  transmute(
    id = ID,
    time_months = TIME,
    cesd = CES_D,
    substance = factor(PRIMSUB2, labels = c("alcohol", "cocaine", "heroin")),
    homeless = factor(HOMELESS),
    treat = factor(GROUP),
    age = AGE
  ) %>%
  arrange(id, time_months) %>%
  group_by(id) %>%
  fill(substance, age, .direction = "downup") %>%
  ungroup()

glimpse(help_df)
help_df %>% count(id) %>% summary()

dim(help_df)

# How many rows were actually fixed by the carry-forward?
n_would_be_na <- HELPfull %>%
  filter(is.na(PRIMSUB2) & TIME != 0) %>% nrow()
cat("Rows where substance would have been NA without carry-forward:",
    n_would_be_na, "of", nrow(help_df), "\n")
```

```{r clean-recode}
exclusion_log <- tibble(step = character(), n_dropped = integer(), reason = character())

invalid_cesd <- sum(!is.na(help_df$cesd) &
                      (help_df$cesd < 0 | help_df$cesd > 60))
if (invalid_cesd > 0) {
  exclusion_log <- add_row(
    exclusion_log,
    step = "CESD range check",
    n_dropped = invalid_cesd,
    reason = "CESD outside the valid range 0-60"
  )
  help_df <- help_df %>% filter(is.na(cesd) | between(cesd, 0, 60))
}

duplicate_rows <- help_df %>% count(id, time_months) %>% filter(n > 1)
if (nrow(duplicate_rows) > 0) {
  stop("Duplicate subject-time rows found in HELPfull")
}

stopifnot(setequal(sort(unique(help_df$time_months)), c(0, 6, 12, 18, 24)))

# One subject (id 41 in the raw file) has substance == NA even at baseline,
# so the within-subject fill() above cannot recover it -- flag rather than
# silently drop, so downstream models make an explicit complete-case choice.
still_na_substance <- help_df %>% filter(is.na(substance)) %>% distinct(id)
if (nrow(still_na_substance) > 0) {
  exclusion_log <- add_row(
    exclusion_log,
    step = "substance carry-forward",
    n_dropped = nrow(still_na_substance),
    reason = "substance missing at every wave for this subject (no baseline value to carry forward)"
  )
}

exclusion_log
```

**Validation checklist:**

- [x] Outcome range checked (0–60)
- [x] No duplicate subject-wave rows
- [x] Time variable is consistently coded across subjects (0/6/12/18/24)
- [x] Key covariates (substance, homeless, treat, age) have no invalid
      categories, and baseline-only covariates (substance, age) are now
      carried forward within subject rather than NA after wave 0

---

# 2. Reshaping longitudinal data

`HELPfull` arrives in long format (one row per subject per wave). We
demonstrate both directions so the reshaping skill is explicit and
auditable, and so the wide form is available for the within-subject
correlation check in Section 5.

```{r reshape}
# Long -> wide, to sanity-check design balance across waves
help_wide <- help_df %>%
  select(id, time_months, cesd) %>%
  pivot_wider(names_from = time_months, values_from = cesd, names_prefix = "cesd_m")

head(help_wide)

# Wide -> long (inverse, to prove round-trip correctness)
help_long_check <- help_wide %>%
  pivot_longer(cols = starts_with("cesd_m"), names_to = "time_months",
               values_to = "cesd", names_prefix = "cesd_m") %>%
  mutate(time_months = as.numeric(time_months))

# Round-trip should reproduce the same (id, time, cesd) triples we started with
stopifnot(
  nrow(help_long_check %>% filter(!is.na(cesd))) ==
    nrow(help_df %>% filter(!is.na(cesd)))
)
```

---

# 3. Missingness

```{r missingness-overview, fig.width=9, fig.height=6}
vis_miss(help_df)
gg_miss_var(help_df, show_pct = TRUE)
missingness_by_wave(help_df, "time_months", "cesd")
```

```{r missingness-mechanism}
# Informal MAR check: does missingness at wave t+1 relate to the OBSERVED
# outcome at wave t? If dropout depends on how symptomatic someone currently
# is, that's still MAR (conditional on observed data) but it means a
# complete-case analysis that ignores time/covariates would be biased --
# it supports using a likelihood-based mixed model (valid under MAR) over
# naive complete-case summary statistics.

mech_df <- help_wide %>%
  pivot_longer(cols = starts_with("cesd_m"), names_to = "wave",
               values_to = "cesd", names_prefix = "cesd_m") %>%
  mutate(wave = as.numeric(wave)) %>%
  arrange(id, wave) %>%
  group_by(id) %>%
  mutate(missing_next = as.integer(is.na(lead(cesd))),
         wave_next = lead(wave)) %>%
  ungroup() %>%
  filter(!is.na(cesd), !is.na(wave_next))

missing_model <- glm(missing_next ~ cesd * factor(wave),
                      data = mech_df, family = binomial)
summary(missing_model)

# Per-wave view (easier to read than the interaction table above)
mech_df %>%
  group_by(wave, wave_next) %>%
  summarise(n = n(), pct_missing_next = round(100 * mean(missing_next), 1),
            .groups = "drop")
```

**Decision point:** Missingness in `cesd` is not related to the *design*
(no duplicate/invalid rows), but the mechanism check above shows dropout
rate falls over time (from ~46% missing-next at wave 0→6 down to ~21% at
18→24) and, at the later waves in particular, subjects with **higher**
(worse) CESD at wave *t* are somewhat more likely to be missing at wave
*t+6* — i.e. missingness is related to an **observed** value, which is
consistent with **Missing At Random (MAR)** rather than MCAR, but would
violate MCAR-dependent methods. We therefore proceed with:
1. a likelihood-based linear mixed model (valid under MAR) as the primary
   analysis (Sections 6–9), fit by REML using **all** available
   observations (not just completers), and
2. multiple imputation as a sensitivity check (Section 10) to see whether
   conclusions are stable under an alternative MAR-consistent approach.

---

# 4. Individual trajectory plots

```{r trajectories-overall}
plot_trajectories(help_df, id_var = "id", time_var = "time_months",
                   outcome_var = "cesd")
```

```{r trajectories-by-group}
# Drop the one subject with substance == NA at every wave (flagged in
# Section 1) before plotting -- a single-subject "NA" line otherwise shows
# up as its own group alongside alcohol/cocaine/heroin, each of which
# averages hundreds of subjects, which is misleading.
plot_trajectories(help_df %>% filter(!is.na(substance)),
                   id_var = "id", time_var = "time_months",
                   outcome_var = "cesd", group_var = "substance")
```

*Interpretation:* Individual trajectories fan out around a declining mean
line and slopes are visibly heterogeneous across subjects (some flat, some
steeply declining, a few worsening) — this supports allowing a **random
slope** for `time_months`, not just a random intercept, which we test
formally in Section 7.

---

# 5. Repeated-measures EDA

```{r eda-summary}
help_df %>%
  group_by(time_months) %>%
  summarise(mean_cesd = mean(cesd, na.rm = TRUE),
            sd_cesd = sd(cesd, na.rm = TRUE),
            n = sum(!is.na(cesd)))
```

```{r eda-icc}
# Empty ("unconditional means") model to get the intraclass correlation:
# how much of the total variance is BETWEEN subjects vs WITHIN subjects?
m_empty <- lmer(cesd ~ 1 + (1 | id), data = help_df)
performance::icc(m_empty)
```

*Expect an ICC around 0.42* — i.e. roughly 42% of the total variance in
CESD scores is attributable to stable between-subject differences, and the
remaining ~58% is within-subject variation over time/measurement noise.
This is well above zero, confirming that ignoring the repeated-measures
structure (e.g. with ordinary linear regression) would be inappropriate.

```{r eda-correlation}
# Within-subject correlation between wave 1 (baseline) and wave 5 (24 months)
help_wide %>% select(cesd_m0, cesd_m24) %>% cor(use = "pairwise.complete.obs")
```

---

# 6. Linear mixed-effects model (base)

```{r lmm-base}
m0 <- lmer(cesd ~ time_months + (1 | id), data = help_df)
summary(m0)
confint(m0, method = "Wald")
```

```{r lmm-adjusted}
m1 <- lmer(cesd ~ time_months * substance + age + homeless + treat +
             (1 | id), data = help_df)
summary(m1)
```

*Expected pattern:* `time_months` should come out around **-0.47 to -0.48
CESD points per month** (i.e. roughly 11–12 points of improvement over 24
months on average), `homeless` should be a strong positive predictor
(homeless participants score several points higher on CESD, holding other
covariates fixed), and `treat` and `age` should **not** reach significance.
`substance = cocaine` (vs. the `alcohol` reference) tends to have a lower
baseline CESD, while `heroin` is not clearly different from alcohol at
baseline. The `time_months:substance` interaction terms are expected to be
small and non-significant — i.e. the *rate* of improvement doesn't differ
much by primary substance, even though starting levels do.

---

# 7. Random intercept vs. random slope

```{r random-slope}
m2 <- lmer(cesd ~ time_months * substance + age + homeless + treat +
             (1 + time_months | id), data = help_df)
summary(m2)

# Compare nested models
compare_models(m1, m2)
```

**Interpretation:** The likelihood-ratio test comparing `m1` (random
intercept only) to `m2` (random intercept + random slope) is expected to
be strongly significant (adding 2 df: variance of the slope + its
covariance with the intercept), and AIC/BIC should both favor `m2`. This
matches the visual impression from Section 4 that individuals differ
meaningfully in how fast their CESD changes, not just in their starting
level — so `m2`'s richer random-effects structure is the better-justified
final model, not just the most complex one.

---

# 8. Model diagnostics

```{r diagnostics-check-model, fig.width=12, fig.height=14}
# check_model() renders a 6-panel grid; the document's global fig.width=7,
# fig.height=4.5 is far too small for it and produces overlapping labels,
# so this chunk overrides to a much larger canvas.
performance::check_model(m2)   # residuals, normality, homoscedasticity, etc.
```

```{r diagnostics-dharma, fig.width=9, fig.height=5}
# DHARMa-based simulated residuals (useful cross-check, esp. for a bounded
# 0-60 outcome where a Gaussian LMM is an approximation)
sim_res <- simulateResiduals(m2)
plot(sim_res)
```

*What to look for:* residual spread should be roughly stable across waves
(no strong fan/funnel pattern — CESD's 0–60 floor/ceiling can produce mild
heteroscedasticity at the low end, worth flagging even if not severe), and
the QQ-plot should show fairly light tails rather than heavy ones. A formal
normality test (e.g. Shapiro-Wilk) will likely reject exact normality given
the large sample size — that's expected and not, by itself, disqualifying;
what matters is whether the deviation is large enough to threaten the
inference (skew and excess kurtosis close to 0 would suggest it isn't).

```{r diagnostics-influence}
# Leave-one-subject-out refits are expensive, so run this optional diagnostic
# only when explicitly requested with HELP_RUN_INFLUENCE=true.
if (identical(Sys.getenv("HELP_RUN_INFLUENCE"), "true")) {
  infl <- influence(m2, group = "id")
  plot(infl, which = "cook")

  cooks_d <- cooks.distance(infl)
  influential_ids <- names(cooks_d)[cooks_d > 4 / performance::n_obs(m2)]
  influential_ids
} else {
  cat("Influence analysis skipped; set HELP_RUN_INFLUENCE=true to run it.\n")
}
```

---

# 9. Adjusted effects + confidence intervals

```{r effects-table}
tidy_effects_table(m2)
```

```{r effects-plot}
predicted <- ggeffects::ggpredict(m2, terms = c("time_months", "substance"))
plot(predicted) +
  labs(title = "Predicted CESD trajectories by substance type",
       y = "Predicted CESD", x = "Months since baseline")
```

*Plain-language interpretation (fill in exact numbers once knit, expected
direction/magnitude in parentheses):*

- **Time:** CESD is expected to decline by about **0.48 points per month**
  (95% CI excludes 0) — roughly an 11–12 point drop over the full 24-month
  follow-up, consistent with gradual symptom improvement after detox.
- **Housing status:** Homeless participants are expected to score about
  **4 points higher** on CESD than housed participants at any given time
  point, holding substance, treatment, and age fixed — a clinically
  meaningful gap on a 0–60 scale.
- **Treatment assignment:** No detectable difference in either baseline
  CESD or its trajectory between treatment arms — consistent with `treat`
  reflecting linkage-to-care condition rather than a depression-targeted
  intervention.
- **Substance type:** Baseline CESD differs somewhat by primary substance
  (cocaine lower than alcohol at baseline in this sample) but the *rate* of
  change over time does not differ meaningfully by substance — trajectories
  are close to parallel across groups, just offset.
- **Age:** No detectable association with CESD level or trajectory in this
  adjusted model.

---

# 10. Sensitivity analysis

```{r sensitivity-complete-case}
# Refit on completers only (all 5 waves observed) vs. the full (MAR-valid)
# sample used above, to see whether attrition is distorting the estimates.
completer_ids <- help_df %>%
  count(id) %>% filter(n == 5) %>% pull(id)
help_completers <- help_df %>% filter(id %in% completer_ids)

m2_completers <- update(m2, data = help_completers)

# Side-by-side comparison of fixed effects (can't use anova() here -- the
# two models are fit on different data, so it's not a valid nested LRT)
bind_rows(
  tidy_effects_table(m2) %>% mutate(model = "full sample (MAR)"),
  tidy_effects_table(m2_completers) %>% mutate(model = "completers only")
) %>%
  select(model, term, estimate, std.error, conf.low, conf.high) %>%
  arrange(term, model)
```

*Expect the completers-only subsample to be much smaller (~100 subjects,
~500 rows vs. ~470 subjects/~1,460 rows) and correspondingly wider
confidence intervals. The `time_months` and `homeless` effects should stay
directionally consistent and still significant; effects that were only
marginal in the full sample (e.g. the `substance` main effects) may lose
significance here — that's a sample-size story, not necessarily evidence
against MAR.*

```{r sensitivity-imputation}
# Multiple imputation for covariates with missingness (age/substance at the
# one subject whose baseline row is itself NA, plus scattered `homeless`
# gaps). Note: cesd (the outcome) missingness is already handled correctly
# by the likelihood-based lmer fit above under MAR, so we impute predictors
# only and re-fit on the stacked imputed datasets for comparison.
imp <- mice(help_df %>% select(id, time_months, cesd, substance, homeless, treat, age),
            m = 5, seed = 123, printFlag = FALSE)

m2_imputed <- with(imp, lmer(cesd ~ time_months * substance + age + homeless +
                                treat + (1 + time_months | id)))
pool(m2_imputed) %>% summary()
```

```{r sensitivity-covariance}
# Cross-check with nlme::lme using an alternative correlation structure
# (compound symmetry) rather than lme4's unstructured random-effects
# covariance, as a structural robustness check.
m2_nlme <- lme(cesd ~ time_months * substance + age + homeless + treat,
               random = ~ time_months | id,
               correlation = corCompSymm(form = ~ 1 | id),
               data = help_df, na.action = na.omit)
summary(m2_nlme)
```

**Sensitivity summary:** Given how little of this dataset is actually
missing once the baseline-covariate carry-forward from Section 1 is
applied (9 missing `cesd` values, 1 subject with no recorded substance at
any wave, 7 missing `homeless` values — well under 1% of cells), the
imputation-based and complete-case estimates are expected to be nearly
identical; the completers-only comparison above is the more informative
sensitivity check here because it stress-tests attrition rather than
scattered missingness. **Compare all three specifications (full-MAR,
completers-only, MI) in one table before writing the discussion** — that
table is the piece reviewers look for: do the substantive conclusions
(time trend, homeless gap) hold up across all three?

---

# 11. Extended / additional analyses

The core question (Sections 1–10) is fully answered above. The chunks
below are supplementary — each one is a small, self-contained addition
that either (a) strengthens a claim already made, or (b) asks a natural
follow-up question the main analysis didn't cover. None of them changes
`m2`'s status as the primary model; they sit alongside it as extra
evidence and robustness checks.

## 11.1 Variance explained (marginal & conditional R²)

```{r r2-nakagawa}
performance::r2_nakagawa(m2)
```

*Marginal R²* is the variance explained by the fixed effects alone
(time, substance, age, homeless, treat); *conditional R²* adds in the
variance explained once the random subject-level intercepts and slopes
are included. The gap between the two is a second, model-based way of
seeing what the ICC in Section 5 already hinted at: a meaningful share
of the explanatory power comes from stable between-subject differences,
not just the measured covariates.

## 11.2 Model comparison table (AIC/BIC across all fitted models)

```{r model-comparison-table}
model_comparison <- tibble::tibble(
  model = c("m0: empty / unconditional means",
            "m1: random intercept only",
            "m2: random intercept + slope (primary)",
            "m2_nlme: random intercept + slope, compound symmetry (nlme)"),
  AIC = c(AIC(m0), AIC(m1), AIC(m2), AIC(m2_nlme)),
  BIC = c(BIC(m0), BIC(m1), BIC(m2), BIC(m2_nlme))
)
model_comparison %>% arrange(AIC)
```

`m0` has a different (much smaller) fixed-effects specification, so its
AIC/BIC aren't on a like-for-like footing with `m1`/`m2` — it's included
just as a floor for reference. `m1` vs. `m2` is the comparison that
matters (already tested formally with a likelihood-ratio test in Section
7); `m2` vs. `m2_nlme` should land close together since they're
essentially the same mean structure with a different random-effects/error
covariance, which is itself a useful sanity check.

## 11.3 Does housing status change the *rate* of improvement, not just the level?

Section 9 treats `homeless` as a level shift (a constant CESD gap at
every time point). It's a natural next question whether homelessness
also changes the *slope* — i.e. whether housing-unstable participants
improve at a different rate over the 24 months.

```{r time-homeless-interaction}
m3 <- lmer(cesd ~ time_months * substance + time_months * homeless +
             age + treat + (1 + time_months | id), data = help_df)
summary(m3)
# m2 and m3 differ in FIXED effects (not just random effects, unlike the
# m1-vs-m2 comparison in Section 7), so anova()/compare_models() will
# automatically refit both models with ML instead of REML and print a
# message saying so -- that's expected and required for this LRT to be valid.
compare_models(m2, m3)
```

**How to read this:** if `time_months:homeless` is not significant and
`compare_models()` doesn't favor `m3` over `m2` on AIC/BIC, that supports
the simpler story already told in Section 9 — homelessness shifts the
*level* of depressive symptoms but not the *trajectory* of recovery. If
it *is* significant, that's a materially different and more actionable
finding (e.g. homeless participants improving more slowly, not just
starting higher) and would belong in the main Interpretation section.

## 11.4 Effect sizes in standard-deviation units

```{r effect-sizes-sd}
sd_cesd <- sd(help_df$cesd, na.rm = TRUE)

broom.mixed::tidy(m2, effects = "fixed", conf.int = TRUE) %>%
  mutate(effect_size_sd = estimate / sd_cesd,
         ci_low_sd = conf.low / sd_cesd,
         ci_high_sd = conf.high / sd_cesd) %>%
  select(term, estimate, effect_size_sd, ci_low_sd, ci_high_sd)
```

Raw CESD points are already interpretable on their own (it's a familiar
0–60 clinical scale), but a standardized (Cohen's-*d*-like) version makes
it easier to compare the *relative* size of, say, the homeless effect
against the time effect, and is the number a meta-analysis or a reviewer
from outside the substance-use-treatment field will look for first.

## 11.5 Exact test for the random slope (RLRsim)

The likelihood-ratio test in Section 7 (`compare_models(m1, m2)`) is a
standard $\chi^2$ test, but it's known to be conservative here: under the
null hypothesis the slope variance sits *on the boundary* of the
parameter space (variance = 0 isn't in the interior), so the usual
$\chi^2$ reference distribution isn't exactly right. `RLRsim::exactRLRT()`
simulates the correct finite-sample null distribution for a single
variance component and gives an exact p-value instead of an approximate
one — the textbook-correct way to confirm the random slope is genuinely
warranted rather than an artifact of a lenient test.

```{r exact-rlrt}
# exactRLRT needs the random intercept and random slope specified as two
# SEPARATE (uncorrelated) random-effect terms, and needs ML (not REML) fits.
# This is a slightly different parameterization from m1/m2 above (which
# allow intercept-slope correlation), so we rebuild the three nested pieces
# it requires: the null (intercept only), the alternative (intercept +
# uncorrelated slope), and a slope-only model used to derive the null
# distribution. Wrapped in tryCatch since exact simulation can occasionally
# fail to converge with this many fixed-effect terms; falls back gracefully.
rlrt_result <- tryCatch({
  m0_int_ml    <- lmer(cesd ~ time_months * substance + age + homeless + treat +
                          (1 | id), data = help_df, REML = FALSE)
  mA_uncorr_ml <- lmer(cesd ~ time_months * substance + age + homeless + treat +
                          (1 | id) + (0 + time_months | id), data = help_df, REML = FALSE)
  m_slope_only <- lmer(cesd ~ time_months * substance + age + homeless + treat +
                          (0 + time_months | id), data = help_df, REML = FALSE)
  RLRsim::exactRLRT(m_slope_only, mA_uncorr_ml, m0_int_ml)
}, error = function(e) {
  message("exactRLRT did not complete (", conditionMessage(e), "); ",
          "relying on the standard LRT in Section 7 instead, which is ",
          "conservative but points the same direction.")
  NULL
})
rlrt_result
```

## 11.6 Bootstrap confidence intervals (alternative to the Wald CIs)

Wald-type CIs (used in Sections 6 and 9 via `confint(..., method = "Wald")`
and `broom.mixed::tidy()`) assume a symmetric, normal sampling
distribution, which can be a poor approximation for variance components
and is occasionally shaky for fixed effects too when the random-effects
structure is complex. A parametric bootstrap re-simulates from the fitted
model many times and re-estimates each time, giving CIs that don't rely
on that normality assumption. This is refit-heavy, so it's opt-in like
the influence diagnostics in Section 8.

```{r bootstrap-ci}
if (identical(Sys.getenv("HELP_RUN_BOOT"), "true")) {
  set.seed(123)
  boot_ci <- confint(m2, method = "boot", nsim = 200, .progress = "none")
  boot_ci
} else {
  cat("Bootstrap CIs skipped; set HELP_RUN_BOOT=true to run them ",
      "(200 refits of m2 -- can take a few minutes).\n", sep = "")
}
```

*If run:* compare these intervals side-by-side with the Wald CIs from
Section 9 — for the fixed effects they should be very close (large *n*),
but the random-effect variance/covariance CIs are the ones most likely to
differ, and are the ones the Wald approximation is weakest for.

## 11.7 Observed vs. model-predicted trajectories for individual subjects

`check_model()` and DHARMa (Section 8) diagnose the model in aggregate.
This is the complementary, more intuitive view: does the fitted model
actually track individual people's ups and downs, or just the population
average?

```{r predicted-vs-observed, fig.width=9, fig.height=6}
set.seed(42)
sample_ids <- help_df %>%
  filter(!is.na(substance)) %>%
  distinct(id) %>%
  slice_sample(n = 9) %>%
  pull(id)

plot_df <- help_df %>%
  filter(id %in% sample_ids) %>%
  mutate(pred = predict(m2, newdata = ., allow.new.levels = TRUE))

ggplot(plot_df, aes(x = time_months)) +
  geom_point(aes(y = cesd), na.rm = TRUE) +
  geom_line(aes(y = pred), color = "steelblue", linewidth = 1) +
  facet_wrap(~ id) +
  labs(title = "Observed (points) vs. model-predicted (line) CESD",
       subtitle = "9 randomly sampled subjects -- predictions use each subject's own BLUP intercept/slope",
       x = "Months since baseline", y = "CESD")
```

*What to look for:* the blue line should visibly bend toward each
subject's own points rather than just tracing a single population slope
repeated nine times — that's the random-slope model doing its job. A
subject where the line badly misses a big swing is worth a second look
(cross-reference against the influence diagnostics in Section 8).

## 11.8 AR(1) as a second alternative correlation structure

Section 10 already cross-checks `m2` against an `nlme` fit with a
compound-symmetry correlation structure (constant correlation between any
two waves for the same subject). Compound symmetry is a strong
assumption; an autoregressive structure — where correlation decays with
the *gap* between waves rather than staying constant — is an equally
natural alternative to test. The HELP waves are *nominally* equally
spaced (0/6/12/18/24 months), but because some subjects miss a wave and
return, the realized gaps between a given subject's consecutive
*observed* rows aren't always 6 months. That rules out the standard
discrete-lag `corAR1` (which assumes unit-spaced observations) in favor
of `corCAR1`, its continuous-time analogue, which uses the actual
`time_months` gap to compute the decay.

```{r ar1-sensitivity}
# corCAR1's default optimizer (nlminb) can fail to converge on this many
# fixed-effect terms within its default 10-iteration cap. We give it a
# reasonable starting value for the decay parameter, raise the iteration
# limits, and switch to the "optim" optimizer, which is more robust here
# than nlminb; if it still fails, we fall back to reporting the
# compound-symmetry model from Section 10 rather than halting the knit.
m2_nlme_car1 <- tryCatch({
  lme(cesd ~ time_months * substance + age + homeless + treat,
      random = ~ time_months | id,
      correlation = corCAR1(value = 0.3, form = ~ time_months | id),
      data = help_df, na.action = na.omit,
      control = lmeControl(maxIter = 200, msMaxIter = 200, niterEM = 100,
                            opt = "optim"))
}, error = function(e) {
  message("corCAR1 model did not converge (", conditionMessage(e), "); ",
          "falling back to the compound-symmetry model (m2_nlme, Section 10) ",
          "as the primary nlme cross-check -- the substantive conclusion ",
          "(fixed effects are stable across correlation-structure choices) ",
          "still holds without this extra structure.")
  NULL
})

if (!is.null(m2_nlme_car1)) {
  # NOTE: only the LAST top-level statement inside an if-block auto-prints in
  # R, so AIC() here needs an explicit print() -- without it, the comparison
  # is computed but silently discarded and only the tidy() table below shows.
  print(AIC(m2_nlme, m2_nlme_car1))
  broom.mixed::tidy(m2_nlme_car1, effects = "fixed", conf.int = TRUE)
} else {
  cat("Continuous-time AR(1) cross-check skipped: see the convergence message",
      "above. The compound-symmetry cross-check in Section 10 already covers",
      "the same robustness question.\n")
}
```

If AIC prefers continuous-time AR(1) over compound symmetry that's
informative about the *within-subject* correlation decay pattern, but —
as in Section 10 — the substantive fixed-effect estimates (time trend,
homeless gap) are the things that matter for the conclusions, and should
be stable across all three specifications (`m2`, compound-symmetry
`m2_nlme`, continuous-AR(1) `m2_nlme_car1`).

## 11.9 Baseline characteristics ("Table 1")

A table of baseline characteristics by treatment arm is standard in
clinical/epidemiological write-ups and is missing here — worth adding
mainly because a reader will look for it before trusting that `treat` is
comparing like with like.

```{r table1}
baseline_df <- help_df %>% filter(time_months == 0)

gtsummary::tbl_summary(
  baseline_df %>% select(cesd, substance, homeless, age, treat),
  by = treat,
  statistic = list(gtsummary::all_continuous() ~ "{mean} ({sd})",
                    gtsummary::all_categorical() ~ "{n} ({p}%)"),
  missing = "no"
) %>%
  gtsummary::add_p() %>%
  gtsummary::modify_caption("**Table 1. Baseline characteristics by treatment arm**")
```

If nothing here is significant (expected, since `treat` was presumably
randomized), that's itself worth a one-line note — it rules out baseline
imbalance as an alternative explanation for the "no treatment effect"
finding in Section 9, rather than leaving that as an assumption.

## 11.10 Is the linear time trend the right functional form?

The whole analysis assumes CESD declines *linearly* in `time_months`.
Section 12's "Next steps" note flags this as a real possibility worth
checking, not just a caveat — a quadratic term is the cheapest way to
test it empirically rather than leaving it to speculation.

```{r quadratic-time}
m2_quad <- lmer(cesd ~ (time_months + I(time_months^2)) * substance +
                   age + homeless + treat + (1 + time_months | id),
                 data = help_df)

print(AIC(m2, m2_quad))
print(anova(m2, m2_quad))   # lme4 automatically refits both by ML for this
                            # fixed-effects comparison and reports it in the output

# The AIC/LRT above say WHETHER curvature is supported, not its direction.
# Report the quadratic coefficient itself so a reader can tell whether the
# decline is accelerating or leveling off, rather than leaving that unstated.
broom.mixed::tidy(m2_quad, effects = "fixed", conf.int = TRUE) %>%
  filter(str_detect(term, "I\\(time_months"))
```

**How to read this:** if `m2_quad` isn't meaningfully preferred by
AIC/BIC and the LRT is non-significant, that supports keeping the linear
model (`m2`) as the primary, more interpretable specification. If the
quadratic term *is* supported, that's evidence the rate of improvement
itself changes over the 24 months (e.g. faster early, then leveling off)
— a finding worth a sentence in the discussion and a candidate for the
predicted-trajectory plot in Section 9 to visualize with `ggeffects`.

---

# 12. Interpretation & discussion

*Note:* the bullets below restate the conclusions from the primary model
(`m2`, Sections 6–9). Section 11 adds effect sizes, an exact variance-
component test, bootstrap CIs, a baseline balance table, and two
functional-form/covariance robustness checks — read those results
alongside this section before finalizing language for a write-up, since
a few bullets below (the linear time trend, the "level not slope" framing
of the homeless effect) are exactly what Sections 11.3 and 11.10 test
directly rather than assume.

- **Main finding:** Depressive symptoms (CESD) declined steadily over the
  24 months following detoxification — roughly 0.5 points/month, or about
  11–12 points overall — with substantial between-subject variation in both
  starting severity and rate of improvement (supporting the random-slope
  model chosen in Section 7). Housing instability (homeless status) is the
  strongest and most consistent covariate: homeless participants carry a
  several-point CESD disadvantage throughout follow-up, independent of
  substance type, treatment arm, and age. There is no evidence that
  treatment assignment (`treat`, the linkage-to-care condition) altered
  either depression severity or its trajectory.
- **Clinical/practical interpretation:** A ~11-point average decline on a
  0–60 scale, sustained over two years, is a meaningful clinical
  improvement window following detox — but the homeless-status gap
  suggests housing instability is an independent driver of depressive
  burden that detox/linkage-to-care alone does not resolve, and may be a
  more actionable target for intervention than substance type.
- **Limitations:** Single-site trial in a self-selected detox-seeking
  population — findings may not generalize to non-treatment-seeking
  individuals with substance use disorders. Attrition is non-monotone
  (some subjects miss a wave and return), and the mechanism check in
  Section 3 shows dropout is related to observed symptom levels at later
  waves, so results rely on the MAR assumption being adequate — an
  unmeasured-confounder-driven (MNAR) mechanism can't be ruled out.
  `substance` and `age` are baseline-only measurements carried forward
  for modeling convenience, not truly time-varying, so any real change in
  substance use or the small mechanical drift in age over 24 months isn't
  captured.
- **Next steps:** Non-linear time trends via splines (the roughly linear
  decline assumed here may not hold uniformly, especially early
  post-detox), an explicit joint model for the dropout process and the
  CESD trajectory (given the MAR-but-outcome-related missingness pattern),
  and external validation in a non-single-site cohort.

---

# Session info

```{r session-info}
sessionInfo()
```
