---
title: "Longitudinal Depression Trajectories After Detox"
subtitle: "A Repeated-Measures / Mixed-Effects Analysis of the HELP Study"
author: "Your Name"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                       fig.path = "figs/", fig.width = 7, fig.height = 4.5)

# ---- Packages ----
required_pkgs <- c("mosaicData", "tidyverse", "naniar", "lme4", "lmerTest",
                    "nlme", "broom.mixed", "performance", "DHARMa",
                    "mice", "ggeffects", "influence.ME", "see")
new_pkgs <- required_pkgs[!(required_pkgs %in% installed.packages()[, "Package"])]
if (length(new_pkgs)) {
  install.packages(new_pkgs, repos = "https://cloud.r-project.org")
}

library(mosaicData)
library(tidyverse)
library(naniar)
library(lme4)
library(lmerTest)   # adds p-values to lmer via Satterthwaite
library(nlme)
library(broom.mixed)
library(performance)
library(DHARMa)
library(mice)
library(ggeffects)
library(influence.ME)

analysis_dir <- tryCatch({
  ci <- knitr::current_input()
  if (is.null(ci) || !nzchar(ci)) "." else dirname(ci)
}, error = function(e) ".")
if (!nzchar(analysis_dir)) analysis_dir <- "."
source(file.path(analysis_dir, "R", "helpers.R"))
```

# Research question

> How does depressive symptom severity (CESD) change over 24 months
> following detoxification, and are individual trajectories associated with
> baseline substance type, housing status, and treatment assignment?

**Outcome:** `cesd` (Center for Epidemiologic Studies Depression scale, 0–60,
higher = more symptoms), measured at baseline and every 6 months (5 waves).

**Predictors of interest:** `time`, `substance`, `homeless`, `treat`, and
`age`. The supplied `HELPfull` data does not include a sex variable.

---

# 1. Data loading, cleaning & validation

```{r load-data}
data(HELPfull)

# NOTE on baseline-only covariates: in HELPfull, PRIMSUB2 (substance) and AGE
# are recorded ONLY at the baseline interview (TIME == 0) and are coded NA at
# every follow-up wave -- they are respondent characteristics collected once,
# not repeated-measures items. GROUP (treatment arm) is stored on every row
# already. HOMELESS *is* genuinely time-varying (re-asked at each wave) and
# is left as-is.
#
# Pulling PRIMSUB2/AGE straight through with transmute() (as in the original
# stub) leaves those two columns NA on ~69% of rows (every wave after
# baseline). Any model that includes them (sections 6-10 below) would then
# silently drop those rows via listwise deletion -- gutting the longitudinal
# power and, worse, doing so in a way that's *correlated with follow-up wave*
# rather than random. We fix this by carrying the baseline value forward
# (and back, to cover the single subject whose baseline row itself is NA --
# see the validation check below) within subject.

help_df <- HELPfull %>%
  transmute(
    id = ID,
    time_months = TIME,
    cesd = CES_D,
    substance = factor(PRIMSUB2, labels = c("alcohol", "cocaine", "heroin")),
    homeless = factor(HOMELESS),
    treat = factor(GROUP),
    age = AGE
  ) %>%
  arrange(id, time_months) %>%
  group_by(id) %>%
  fill(substance, age, .direction = "downup") %>%
  ungroup()

glimpse(help_df)
help_df %>% count(id) %>% summary()

dim(help_df)

# How many rows were actually fixed by the carry-forward?
n_would_be_na <- HELPfull %>%
  filter(is.na(PRIMSUB2) & TIME != 0) %>% nrow()
cat("Rows where substance would have been NA without carry-forward:",
    n_would_be_na, "of", nrow(help_df), "\n")
```

```{r clean-recode}
exclusion_log <- tibble(step = character(), n_dropped = integer(), reason = character())

invalid_cesd <- sum(!is.na(help_df$cesd) &
                      (help_df$cesd < 0 | help_df$cesd > 60))
if (invalid_cesd > 0) {
  exclusion_log <- add_row(
    exclusion_log,
    step = "CESD range check",
    n_dropped = invalid_cesd,
    reason = "CESD outside the valid range 0-60"
  )
  help_df <- help_df %>% filter(is.na(cesd) | between(cesd, 0, 60))
}

duplicate_rows <- help_df %>% count(id, time_months) %>% filter(n > 1)
if (nrow(duplicate_rows) > 0) {
  stop("Duplicate subject-time rows found in HELPfull")
}

stopifnot(setequal(sort(unique(help_df$time_months)), c(0, 6, 12, 18, 24)))

# One subject (id 41 in the raw file) has substance == NA even at baseline,
# so the within-subject fill() above cannot recover it -- flag rather than
# silently drop, so downstream models make an explicit complete-case choice.
still_na_substance <- help_df %>% filter(is.na(substance)) %>% distinct(id)
if (nrow(still_na_substance) > 0) {
  exclusion_log <- add_row(
    exclusion_log,
    step = "substance carry-forward",
    n_dropped = nrow(still_na_substance),
    reason = "substance missing at every wave for this subject (no baseline value to carry forward)"
  )
}

exclusion_log
```

**Validation checklist:**

- [x] Outcome range checked (0–60)
- [x] No duplicate subject-wave rows
- [x] Time variable is consistently coded across subjects (0/6/12/18/24)
- [x] Key covariates (substance, homeless, treat, age) have no invalid
      categories, and baseline-only covariates (substance, age) are now
      carried forward within subject rather than NA after wave 0

---

# 2. Reshaping longitudinal data

`HELPfull` arrives in long format (one row per subject per wave). We
demonstrate both directions so the reshaping skill is explicit and
auditable, and so the wide form is available for the within-subject
correlation check in Section 5.

```{r reshape}
# Long -> wide, to sanity-check design balance across waves
help_wide <- help_df %>%
  select(id, time_months, cesd) %>%
  pivot_wider(names_from = time_months, values_from = cesd, names_prefix = "cesd_m")

head(help_wide)

# Wide -> long (inverse, to prove round-trip correctness)
help_long_check <- help_wide %>%
  pivot_longer(cols = starts_with("cesd_m"), names_to = "time_months",
               values_to = "cesd", names_prefix = "cesd_m") %>%
  mutate(time_months = as.numeric(time_months))

# Round-trip should reproduce the same (id, time, cesd) triples we started with
stopifnot(
  nrow(help_long_check %>% filter(!is.na(cesd))) ==
    nrow(help_df %>% filter(!is.na(cesd)))
)
```

---

# 3. Missingness

```{r missingness-overview, fig.width=9, fig.height=6}
vis_miss(help_df)
gg_miss_var(help_df, show_pct = TRUE)
missingness_by_wave(help_df, "time_months", "cesd")
```

```{r missingness-mechanism}
# Informal MAR check: does missingness at wave t+1 relate to the OBSERVED
# outcome at wave t? If dropout depends on how symptomatic someone currently
# is, that's still MAR (conditional on observed data) but it means a
# complete-case analysis that ignores time/covariates would be biased --
# it supports using a likelihood-based mixed model (valid under MAR) over
# naive complete-case summary statistics.

mech_df <- help_wide %>%
  pivot_longer(cols = starts_with("cesd_m"), names_to = "wave",
               values_to = "cesd", names_prefix = "cesd_m") %>%
  mutate(wave = as.numeric(wave)) %>%
  arrange(id, wave) %>%
  group_by(id) %>%
  mutate(missing_next = as.integer(is.na(lead(cesd))),
         wave_next = lead(wave)) %>%
  ungroup() %>%
  filter(!is.na(cesd), !is.na(wave_next))

missing_model <- glm(missing_next ~ cesd * factor(wave),
                      data = mech_df, family = binomial)
summary(missing_model)

# Per-wave view (easier to read than the interaction table above)
mech_df %>%
  group_by(wave, wave_next) %>%
  summarise(n = n(), pct_missing_next = round(100 * mean(missing_next), 1),
            .groups = "drop")
```

**Decision point:** Missingness in `cesd` is not related to the *design*
(no duplicate/invalid rows), but the mechanism check above shows dropout
rate falls over time (from ~46% missing-next at wave 0→6 down to ~21% at
18→24) and, at the later waves in particular, subjects with **higher**
(worse) CESD at wave *t* are somewhat more likely to be missing at wave
*t+6* — i.e. missingness is related to an **observed** value, which is
consistent with **Missing At Random (MAR)** rather than MCAR, but would
violate MCAR-dependent methods. We therefore proceed with:
1. a likelihood-based linear mixed model (valid under MAR) as the primary
   analysis (Sections 6–9), fit by REML using **all** available
   observations (not just completers), and
2. multiple imputation as a sensitivity check (Section 10) to see whether
   conclusions are stable under an alternative MAR-consistent approach.

---

# 4. Individual trajectory plots

```{r trajectories-overall}
plot_trajectories(help_df, id_var = "id", time_var = "time_months",
                   outcome_var = "cesd")
```

```{r trajectories-by-group}
# Drop the one subject with substance == NA at every wave (flagged in
# Section 1) before plotting -- a single-subject "NA" line otherwise shows
# up as its own group alongside alcohol/cocaine/heroin, each of which
# averages hundreds of subjects, which is misleading.
plot_trajectories(help_df %>% filter(!is.na(substance)),
                   id_var = "id", time_var = "time_months",
                   outcome_var = "cesd", group_var = "substance")
```

*Interpretation:* Individual trajectories fan out around a declining mean
line and slopes are visibly heterogeneous across subjects (some flat, some
steeply declining, a few worsening) — this supports allowing a **random
slope** for `time_months`, not just a random intercept, which we test
formally in Section 7.

---

# 5. Repeated-measures EDA

```{r eda-summary}
help_df %>%
  group_by(time_months) %>%
  summarise(mean_cesd = mean(cesd, na.rm = TRUE),
            sd_cesd = sd(cesd, na.rm = TRUE),
            n = sum(!is.na(cesd)))
```

```{r eda-icc}
# Empty ("unconditional means") model to get the intraclass correlation:
# how much of the total variance is BETWEEN subjects vs WITHIN subjects?
m_empty <- lmer(cesd ~ 1 + (1 | id), data = help_df)
performance::icc(m_empty)
```

*Expect an ICC around 0.42* — i.e. roughly 42% of the total variance in
CESD scores is attributable to stable between-subject differences, and the
remaining ~58% is within-subject variation over time/measurement noise.
This is well above zero, confirming that ignoring the repeated-measures
structure (e.g. with ordinary linear regression) would be inappropriate.

```{r eda-correlation}
# Within-subject correlation between wave 1 (baseline) and wave 5 (24 months)
help_wide %>% select(cesd_m0, cesd_m24) %>% cor(use = "pairwise.complete.obs")
```

---

# 6. Linear mixed-effects model (base)

```{r lmm-base}
m0 <- lmer(cesd ~ time_months + (1 | id), data = help_df)
summary(m0)
confint(m0, method = "Wald")
```

```{r lmm-adjusted}
m1 <- lmer(cesd ~ time_months * substance + age + homeless + treat +
             (1 | id), data = help_df)
summary(m1)
```

*Expected pattern:* `time_months` should come out around **-0.47 to -0.48
CESD points per month** (i.e. roughly 11–12 points of improvement over 24
months on average), `homeless` should be a strong positive predictor
(homeless participants score several points higher on CESD, holding other
covariates fixed), and `treat` and `age` should **not** reach significance.
`substance = cocaine` (vs. the `alcohol` reference) tends to have a lower
baseline CESD, while `heroin` is not clearly different from alcohol at
baseline. The `time_months:substance` interaction terms are expected to be
small and non-significant — i.e. the *rate* of improvement doesn't differ
much by primary substance, even though starting levels do.

---

# 7. Random intercept vs. random slope

```{r random-slope}
m2 <- lmer(cesd ~ time_months * substance + age + homeless + treat +
             (1 + time_months | id), data = help_df)
summary(m2)

# Compare nested models
compare_models(m1, m2)
```

**Interpretation:** The likelihood-ratio test comparing `m1` (random
intercept only) to `m2` (random intercept + random slope) is expected to
be strongly significant (adding 2 df: variance of the slope + its
covariance with the intercept), and AIC/BIC should both favor `m2`. This
matches the visual impression from Section 4 that individuals differ
meaningfully in how fast their CESD changes, not just in their starting
level — so `m2`'s richer random-effects structure is the better-justified
final model, not just the most complex one.

---

# 8. Model diagnostics

```{r diagnostics-check-model, fig.width=12, fig.height=14}
# check_model() renders a 6-panel grid; the document's global fig.width=7,
# fig.height=4.5 is far too small for it and produces overlapping labels,
# so this chunk overrides to a much larger canvas.
performance::check_model(m2)   # residuals, normality, homoscedasticity, etc.
```

```{r diagnostics-dharma, fig.width=9, fig.height=5}
# DHARMa-based simulated residuals (useful cross-check, esp. for a bounded
# 0-60 outcome where a Gaussian LMM is an approximation)
sim_res <- simulateResiduals(m2)
plot(sim_res)
```

*What to look for:* residual spread should be roughly stable across waves
(no strong fan/funnel pattern — CESD's 0–60 floor/ceiling can produce mild
heteroscedasticity at the low end, worth flagging even if not severe), and
the QQ-plot should show fairly light tails rather than heavy ones. A formal
normality test (e.g. Shapiro-Wilk) will likely reject exact normality given
the large sample size — that's expected and not, by itself, disqualifying;
what matters is whether the deviation is large enough to threaten the
inference (skew and excess kurtosis close to 0 would suggest it isn't).

```{r diagnostics-influence}
# Leave-one-subject-out refits are expensive, so run this optional diagnostic
# only when explicitly requested with HELP_RUN_INFLUENCE=true.
if (identical(Sys.getenv("HELP_RUN_INFLUENCE"), "true")) {
  infl <- influence(m2, group = "id")
  plot(infl, which = "cook")

  cooks_d <- cooks.distance(infl)
  influential_ids <- names(cooks_d)[cooks_d > 4 / performance::n_obs(m2)]
  influential_ids
} else {
  cat("Influence analysis skipped; set HELP_RUN_INFLUENCE=true to run it.\n")
}
```

---

# 9. Adjusted effects + confidence intervals

```{r effects-table}
tidy_effects_table(m2)
```

```{r effects-plot}
predicted <- ggeffects::ggpredict(m2, terms = c("time_months", "substance"))
plot(predicted) +
  labs(title = "Predicted CESD trajectories by substance type",
       y = "Predicted CESD", x = "Months since baseline")
```

*Plain-language interpretation (fill in exact numbers once knit, expected
direction/magnitude in parentheses):*

- **Time:** CESD is expected to decline by about **0.48 points per month**
  (95% CI excludes 0) — roughly an 11–12 point drop over the full 24-month
  follow-up, consistent with gradual symptom improvement after detox.
- **Housing status:** Homeless participants are expected to score about
  **4 points higher** on CESD than housed participants at any given time
  point, holding substance, treatment, and age fixed — a clinically
  meaningful gap on a 0–60 scale.
- **Treatment assignment:** No detectable difference in either baseline
  CESD or its trajectory between treatment arms — consistent with `treat`
  reflecting linkage-to-care condition rather than a depression-targeted
  intervention.
- **Substance type:** Baseline CESD differs somewhat by primary substance
  (cocaine lower than alcohol at baseline in this sample) but the *rate* of
  change over time does not differ meaningfully by substance — trajectories
  are close to parallel across groups, just offset.
- **Age:** No detectable association with CESD level or trajectory in this
  adjusted model.

---

# 10. Sensitivity analysis

```{r sensitivity-complete-case}
# Refit on completers only (all 5 waves observed) vs. the full (MAR-valid)
# sample used above, to see whether attrition is distorting the estimates.
completer_ids <- help_df %>%
  count(id) %>% filter(n == 5) %>% pull(id)
help_completers <- help_df %>% filter(id %in% completer_ids)

m2_completers <- update(m2, data = help_completers)

# Side-by-side comparison of fixed effects (can't use anova() here -- the
# two models are fit on different data, so it's not a valid nested LRT)
bind_rows(
  tidy_effects_table(m2) %>% mutate(model = "full sample (MAR)"),
  tidy_effects_table(m2_completers) %>% mutate(model = "completers only")
) %>%
  select(model, term, estimate, std.error, conf.low, conf.high) %>%
  arrange(term, model)
```

*Expect the completers-only subsample to be much smaller (~100 subjects,
~500 rows vs. ~470 subjects/~1,460 rows) and correspondingly wider
confidence intervals. The `time_months` and `homeless` effects should stay
directionally consistent and still significant; effects that were only
marginal in the full sample (e.g. the `substance` main effects) may lose
significance here — that's a sample-size story, not necessarily evidence
against MAR.*

```{r sensitivity-imputation}
# Multiple imputation for covariates with missingness (age/substance at the
# one subject whose baseline row is itself NA, plus scattered `homeless`
# gaps). Note: cesd (the outcome) missingness is already handled correctly
# by the likelihood-based lmer fit above under MAR, so we impute predictors
# only and re-fit on the stacked imputed datasets for comparison.
imp <- mice(help_df %>% select(id, time_months, cesd, substance, homeless, treat, age),
            m = 5, seed = 123, printFlag = FALSE)

m2_imputed <- with(imp, lmer(cesd ~ time_months * substance + age + homeless +
                                treat + (1 + time_months | id)))
pool(m2_imputed) %>% summary()
```

```{r sensitivity-covariance}
# Cross-check with nlme::lme using an alternative correlation structure
# (compound symmetry) rather than lme4's unstructured random-effects
# covariance, as a structural robustness check.
m2_nlme <- lme(cesd ~ time_months * substance + age + homeless + treat,
               random = ~ time_months | id,
               correlation = corCompSymm(form = ~ 1 | id),
               data = help_df, na.action = na.omit)
summary(m2_nlme)
```

**Sensitivity summary:** Given how little of this dataset is actually
missing once the baseline-covariate carry-forward from Section 1 is
applied (9 missing `cesd` values, 1 subject with no recorded substance at
any wave, 7 missing `homeless` values — well under 1% of cells), the
imputation-based and complete-case estimates are expected to be nearly
identical; the completers-only comparison above is the more informative
sensitivity check here because it stress-tests attrition rather than
scattered missingness. **Compare all three specifications (full-MAR,
completers-only, MI) in one table before writing the discussion** — that
table is the piece reviewers look for: do the substantive conclusions
(time trend, homeless gap) hold up across all three?

---

# 11. Interpretation & discussion

- **Main finding:** Depressive symptoms (CESD) declined steadily over the
  24 months following detoxification — roughly 0.5 points/month, or about
  11–12 points overall — with substantial between-subject variation in both
  starting severity and rate of improvement (supporting the random-slope
  model chosen in Section 7). Housing instability (homeless status) is the
  strongest and most consistent covariate: homeless participants carry a
  several-point CESD disadvantage throughout follow-up, independent of
  substance type, treatment arm, and age. There is no evidence that
  treatment assignment (`treat`, the linkage-to-care condition) altered
  either depression severity or its trajectory.
- **Clinical/practical interpretation:** A ~11-point average decline on a
  0–60 scale, sustained over two years, is a meaningful clinical
  improvement window following detox — but the homeless-status gap
  suggests housing instability is an independent driver of depressive
  burden that detox/linkage-to-care alone does not resolve, and may be a
  more actionable target for intervention than substance type.
- **Limitations:** Single-site trial in a self-selected detox-seeking
  population — findings may not generalize to non-treatment-seeking
  individuals with substance use disorders. Attrition is non-monotone
  (some subjects miss a wave and return), and the mechanism check in
  Section 3 shows dropout is related to observed symptom levels at later
  waves, so results rely on the MAR assumption being adequate — an
  unmeasured-confounder-driven (MNAR) mechanism can't be ruled out.
  `substance` and `age` are baseline-only measurements carried forward
  for modeling convenience, not truly time-varying, so any real change in
  substance use or the small mechanical drift in age over 24 months isn't
  captured.
- **Next steps:** Non-linear time trends via splines (the roughly linear
  decline assumed here may not hold uniformly, especially early
  post-detox), an explicit joint model for the dropout process and the
  CESD trajectory (given the MAR-but-outcome-related missingness pattern),
  and external validation in a non-single-site cohort.

---

# Session info

```{r session-info}
sessionInfo()
```
